How to select predictive models for decision-making or causal inference?

### 2024-08-29 response_to_reviewer.md

https://www.research-collection.ethz.ch/handle/20.500.11850/650703

https://aisel.aisnet.org/pacis2022/181/

or some other that can be easily found in literature.

> Thanks for the references. I am in favor of adding some of them to the manuscript such as [Chernozhukov et al., 2024](https://causalml-book.org/) or [Kaddour et al., 2022](https://arxiv.org/pdf/2206.15475) which clearly distangle the different subfields at the intersection of causal inference and machine learning. I have added them at the beginning of the manuscript before focusing on the model selection problem for cate.

@@

Overall: I am no fully convinced that R-risk is always the best.  The answer could be dependent on other factors and I hope the authors would explore more scenarios.

**TODO**

Another question: do the differences between risks in experiments follow known theoretical results?  For example, it is well known (e.g. https://www.semanticscholar.org/reader/112aba9c62f98b9684693f4c20c30274bc9eb1dc or [Rudas ans jaroszewisz, 2017](https://link.springer.com/content/pdf/10.1007/s10618-018-0576-8.pdf)) that the tau risk performs poorly.

> The high variability of the U-learner has already been noted by [Nie and Wager, 2017](https://arxiv.org/pdf/1712.04912). This instability is naturally found in the U-risk. For the other risk, we are not aware of any theoretical results justifying differences between risks. 
> Empirically, the differences between risks have been highlighted in other papers such as [Schuler, et al., 2018](https://arxiv.org/pdf/1804.05146) for quadratic and stepwise data generation processes or [Curth et Van der Schaar, 2023](https://proceedings.mlr.press/v202/curth23b/curth23b.pdf) exploring more diverse outcome generation processes focusing on CATE sparsity (as a proxy of its complexity) based on a semi-simulated dataset for the covariate (ACIC2016). They both underline good performances for the R-risk (see Table.1 for Schuler and Fig. 3 for Curth). A remark that we made in our manuscript is that the results of Schuler are convincing for randomization setups but not for observational (non-randomized) setups. Curth's results are more convincing for observational setups but did not explore the overlap assumption.

### 2024-08-31 response_to_reviewer.md

Other reference on g-computation to consider in the study can be for example:

-       [Le Borgne et al. 2021](https://www.nature.com/articles/s41598-021-81110-0)
-       https://www.sciencedirect.com/science/article/pii/S0167947306002076
-       https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-023-01835-6
-       https://pubmed.ncbi.nlm.nih.gov/34861799/

I understand that Eq. (2) and in Eq. (3) consider both treated and non-treated sample (or patient?), but those sums can be also equal to 0, even if their single terms are different from 0. Moreover, it is not clear if, and how Eq. (3) is related to Eq. (2): is it a variant? Is Eq. (3) derived from Eq. (2) under special assumptions?

> Eq. (3) is the definition of the estimator of the ATE based on a model of the outcome. It is sometimes called g-estimation. It  is unbiased under under consistency (Assumption. 3) and unconfoundedness (Assumption 1). To prove that, take the expectation of Eq. (3).  Then apply  Eq. (1) which is true under consistency (Assumption 3) and unconfoundedness (Assumption 1). Then by the law of total expectancy, the conditionnality on X disappear and we recover Eq. (2).
> G-estimation itself does not control error on both populations as highlighted in Figure 1. Specific inductive biases of ML models are able to favor such regularization(see for example [Johannson et al., 2022](https://www.jmlr.org/papers/volume23/19-511/19-511.pdf)). Selection procedures, more elaborated than RMSE are needed to favor such models. I added this last explanation in the method (par. Model-selection risks, oracle and feasible) trying to clarify the link between the hint on error control on both population, g-estimation and the selection procedure which is our main focus.

@@

Furthermore, the Robinson decomposition (R-decomposition) should be better discussed and its role in the context of causal inference should be better commented.
R-decomposition introduces two quantities, the conditional mean outcome and the probability to be treated (known as propensity score). What is the role of the propensity score? How is it calculated? Is it calculated simply by a frequentist approach (number fo treated/total number of patients?

> G-computation is one choice of decomposition of the CATE estimation into a sequence of regressions, where the only statistical estimate is the response function µa(x) [49 , chapter 15]. Other choices of decomposition exist, such as the R-decomposition [57], the Double robust estimation, the X-learner... The R-decomposition is one way of identifying the causal quantity of interest which introduce two statistical quantity that have to be modelled: the mean outcome from the treatment effect and the propensity score. Any statistical estimator can be chosen to model these quantities. The frequentist approach is one (naive) such estimator. In practice, researchers often estimate the propensity score with a logistic regression, but other methods such as Boosted trees, Random Forests, or Neural Networks can be used. A good ps estimator should be well calibrated but this question is out of scope for this paper. We briefly mentioned in the discussion this research question on selecting a good estimator for the propensity score.

@@

https://www.research-collection.ethz.ch/handle/20.500.11850/650703

https://aisel.aisnet.org/pacis2022/181/

or some other that can be easily found in literature.

> Thanks for the references. I added two references to Causal ML to the manuscript such as [Chernozhukov et al., 2024](https://causalml-book.org/) or [Kaddour et al., 2022](https://arxiv.org/pdf/2206.15475). Both of them cover most of the subfields at the intersection of causal inference and machine learning. I have added them at the beginning of the manuscript before stating our focus on model selection for CATE.

### 2024-08-31 how_to_select_latex.tex
answer to R1

    \begin{keywords}
        Model Selection, Predictive model, Treatment Effect, G-formula, Machine Learning
    \end{keywords}

@@

Indeed, concluding on the effect of a treatment is a difficult
causal-inference task, as it can be easily compromised by confounding:
spurious associations between treatment
allocation and baseline health, \eg~only prescribing a drug to mild cases
\cite{hernan_causal_2020,vanderweele2019principles}.
Predictive modeling is linked to causal inference theory by the concept of
\emph{outcome models} (or G-computation,
G-formula \cite{robins_role_1986}, Q-model
\cite{snowden_implementation_2011}, conditional mean regression
\cite{wendling_comparing_2018}).
% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
% alleviate this concern. Yet most machine-learning models are trained on
% \emph{observational} data, close to real-word practice
% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
Medical statistics and epidemiology have mostly used other
causal-inference methods, modeling treatment assignment
with propensity scores \cite{rosenbaum_central_1983,austin_moving_2015,casucci2018estimating,grose_use_2020}. Outcome modeling brings
the benefit of going
beyond average effects, estimating individualized or conditional average
treatment effects (CATE), central to precision medicine.
%
For this purpose, such methods are also invaluable for randomized trials
\cite{su2018random,lamont2018identification,hoogland2021tutorial}.

@@

%\idea{There is a variety of models for g-estimation}
% There is an explosion of outcome modeling or machine learning methods, either 
% developed for
% For instance, many deep-learning methods have been developed for medical
% image analysis \cite{shen2017deep,monshi2020deep}.
Outcome-modeling methods, even when specifically designed for causal
inference, are numerous: Bayesian Additive Regression Trees
\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
\cite{powers_methods_2018}, causal multivariate adaptive regression
splines \cite{powers_methods_2018}, random forests
\cite{wager_estimation_2018, athey_generalized_2019},
Meta-learners \cite{kunzel_metalearners_2019}, R-learners
\cite{nie_quasioracle_2017}, Doubly robust estimation
\cite{chernozhukov_double_2018}...
%\idea{The variety of estimators calls for model selection procedures}
The wide variety of methods raises the problem
of selecting between different estimators based on the data at hand. %TODO: might be removed since repeated two sentences below
%
Indeed, estimates of treatment effects can vary markedly across different
predictive models \cite{fang2019applying,dorie_automated_2019} (illustration in
Appendix \ref{apd:toy_example:acic_2016_ate_variability}).

@@

We study \textit{model selection procedures} in practical settings,
without theoretical assumptions often made in statistical literature such
as \textit{infinite} data or \textit{well-specified} models (Appendix
\ref{apd:prior_work}). Asymptotic causal-inference theory recommends complex
risks, but a practical question is whether model-selection procedures, that rely
on data split, can estimate these risks reliably enough. Indeed, these
risks come with
more quantities to estimate, which may bring additional variance, leading to
worse model selection.

@@

This toy example illustrates that the classic minimum Mean Square Error
criterion is not suited to choosing a model among candidate
estimators for causal inference.

@@

Should we know the two potential outcomes for a given $X$,
we could compute the
difference between them, which gives the causal effect of the treatment.
%
These two potential outcomes can be estimated from observed data:
the consistency \ref{assumption:consistency} and unconfoundedness
\ref{assumption:ignorability} assumptions imply the following equality:
\begin{equation}\label{eq:mu_identification}
    \mathbb E_{Y(a) \sim \mathcal{D^{\star}}} [Y(a)|X=x] = \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
\end{equation}
On the left, the expectation is taken on the counterfactual unobserved
distribution. On the right, the expectation is taken on the factual observed
distribution conditionally on the treatment. For the rest of the
paper, the expectations will always be taken on the factual observed
distribution $\mathcal{D}$. This identification leads to outcome based estimators (ie.
g-computation estimators \cite{snowden_implementation_2011}):
\begin{eqnarray}
    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)]
    \notag
    \\
    = &\mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
    \label{eq:tau_population}
\end{eqnarray}
This equation builds on the conditional expectation of the outcome given the treatment $\mathbb E_{\sim \mathcal{D}}[Y|A]$. Outcome based methods target this quantity conditionally on the covariates, called \emph{response function}:
\begin{flalign*}
    \text{Response function}
     &  &
    \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
     &  &
\end{flalign*}

@@

%TODO: not well located here. The reader does not know why we introduce this decomoositon her.
\paragraph{Robinson decomposition}
The \emph{R-decomposition}
of the outcome model\cite{robinson_rootnconsistent_1988} links the counterfactual outcomes $y(a)$ to the CATE by introducing two quantities, the conditional mean outcome
and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}).:
\begin{flalign}
    \text{Conditional mean outcome} &                    & m(x) \myeq \; & \mathbb E_{Y \sim
    \mathcal{D}} [Y|X=x]            &                    &
    \label{def:m}
    \\
    \text{Propensity score}         &                    &
    e(x) \myeq \;                   & \mathbb P[A=1|X=x]
    \label{def:propensity_score}
\end{flalign}
the outcome can be written
\begin{flalign}\label{eq:r_decomposition}
    \text{R-decomposition}            &  & y(a) = m(x) + \big( a - e(x) \big)
    \tau(x) + \varepsilon(x; a)\notag &  &
    \\\ &&\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0&&
\end{flalign}
$m$ and $e$ are often called
\emph{nuisances} \cite{chernozhukov_double_2018}; they are unknown and are estimated in practice thanks to statistical models.

@@

We formalize model selection for causal estimation. Thanks to the g-formula
identification (\autoref{eq:mu_identification}), a given outcome model $f: \mathcal X
    \times \mathcal A \rightarrow \mathcal{Y}$ --learned from data or built from
domain knowledge-- induces feasible estimates of the ATE and CATE (eqs
\ref{eq:ate_estimate} and \ref{eq:cate_estimate}), $\hat \tau_{f}$ and $\hat \tau_{f}(x)$.
%
Let $\mathcal F=\{f: \mathcal X \times \mathcal A \rightarrow \mathcal{Y}\}$ be
a family of such estimators. Our goal is to select the best candidate in this
family for the observed dataset $O$ using a risk
$\ell$:
\begin{equation}
    f^*_{\ell} = \argmin_{f \in \mathcal{F}} \ell(f, O)
    \label{eq:causal_model_selection}
\end{equation}

@@

Using eight simulations setups from \cite{powers_methods_2018}, where
the oracle CATE is known, \cite{schuler_comparison_2018} compare four
causal risks, concluding that for CATE estimation the best
model-selection risk is the so-called $R\text{-risk}$
\cite{nie_quasioracle_2017} --def.\,\ref{def:r_risk}, below. Their
empirical results are clear for randomized treatment allocation but less
convincing for observational settings where both simple Mean Squared
Error --MSE, $\mu\text{-risk}(f)$ def.\,\ref{def:mu_risk}-- and
reweighted MSE --$\mu\text{-risk}_{IPW}$ def.\,\ref{def:mu_ipw_risk}--
appear to perform better than $R\text{-risk}$ on half of the simulations.
Another work \cite{alaa_validating_2019} studied empirically both MSE and
reweighted MSE risks on the semi-synthetic ACIC 2016 datasets
\cite{dorie_automated_2019}, but did not include the $R\text{-risk}$. We complete these
prior empirical work by studying a wider variety of data generative
processes and varying the influence of overlap, an important parameter of
the data generation process which makes a given causal metric appropriate
\cite{damour_overlap_2020}. We also study how to best adapt
cross-validation procedures to causal metrics which themselves come with
models to estimate.

>>>>

    \begin{keywords}
        Model Selection, Predictive model, Treatment Effect, G-computation, Machine Learning
    \end{keywords}

@@

Indeed, concluding on the effect of a treatment is a difficult
causal-inference task, as it can be easily compromised by confounding:
spurious associations between treatment
allocation and baseline health, \eg~only prescribing a drug to mild cases
\cite{hernan_causal_2020,vanderweele2019principles}.
Predictive modeling is linked to causal inference theory by the concept of
\emph{outcome models} (or g-computation, g-estimation,
g-formula \cite{robins_role_1986}, Q-model
\cite{snowden_implementation_2011}, conditional mean regression
\cite{wendling_comparing_2018}).
% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
% alleviate this concern. Yet most machine-learning models are trained on
% \emph{observational} data, close to real-word practice
% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
Medical statistics and epidemiology have mostly used other
causal-inference methods, modeling treatment assignment
with propensity scores \cite{rosenbaum_central_1983,austin_moving_2015,casucci2018estimating,grose_use_2020}. Outcome modeling brings
the benefit of going
beyond average effects, estimating individualized or conditional average
treatment effects (CATE), central to precision medicine.
%
For this purpose, such methods are also invaluable for randomized trials
\cite{su2018random,lamont2018identification,hoogland2021tutorial}.

@@

%\idea{There is a variety of models for g-computation}
% There is an explosion of outcome modeling or machine learning methods, either 
% developed for
% For instance, many deep-learning methods have been developed for medical
% image analysis \cite{shen2017deep,monshi2020deep}.
Outcome-modeling methods, even when specifically designed for causal
inference, are numerous: Bayesian Additive Regression Trees
\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
\cite{powers_methods_2018}, causal multivariate adaptive regression
splines \cite{powers_methods_2018}, random forests
\cite{wager_estimation_2018, athey_generalized_2019},
Meta-learners \cite{kunzel_metalearners_2019}, R-learners
\cite{nie_quasioracle_2017}, Doubly robust estimation
\cite{chernozhukov_double_2018}...
%\idea{The variety of estimators calls for model selection procedures}
The wide variety of methods raises the problem
of selecting between different estimators based on the data at hand. %TODO: might be removed since repeated two sentences below
%
Indeed, estimates of treatment effects can vary markedly across different
predictive models \cite{fang2019applying,dorie_automated_2019,leborgne2021g,ren2023comparing} (illustration in
Appendix \ref{apd:toy_example:acic_2016_ate_variability}).

@@

The intersection between machine learning and causal inference is growing rapidly
\cite{kaddour2022causal,chernozhukov2024applied}. We focus on \textit{model selection procedures} in practical settings,
without theoretical assumptions often made in statistical literature such
as \textit{infinite} data or \textit{well-specified} models (Appendix
\ref{apd:prior_work}). Asymptotic causal-inference theory recommends complex
risks, but a practical question is whether model-selection procedures, that rely
on data split, can estimate these risks reliably enough. Indeed, these
risks come with
more quantities to estimate, which may bring additional variance, leading to
worse model selection.

@@

This toy example illustrates that the classic minimum mean squared error (MSE)
criterion is not suited to choosing a model among candidate
estimators for causal inference.

@@

Should we know the two potential outcomes for a given $X$,
we could compute the
difference between them, which gives the causal effect of the treatment.
%
These two potential outcomes can be estimated from observed data:
the consistency \ref{assumption:consistency} and unconfoundedness
\ref{assumption:ignorability} assumptions imply the following equality, linking the target potential quantity to the observed data:
\begin{equation}\label{eq:mu_identification}
    \mathbb E_{Y(a) \sim \mathcal{D^{\star}}} [Y(a)|X=x] = \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
\end{equation}
On the left, the expectation is taken on the counterfactual unobserved
distribution. On the right, the expectation is taken on the factual observed
distribution conditionally on the treatment. For the rest of the
paper, the expectations will always be taken on the factual observed
distribution $\mathcal{D}$. This identification leads to outcome based estimators (ie.
g-computation estimators \cite{snowden_implementation_2011}):
\begin{eqnarray}
    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)]
    \notag
    \\
    = &\mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
    \label{eq:tau_population}
\end{eqnarray}
This equation builds on the conditional expectation of the outcome given the treatment $\mathbb E_{\sim \mathcal{D}}[Y|A]$. Outcome based methods target this quantity conditionally on the covariates, called \emph{response function}:
\begin{flalign*}
    \text{Response function}
     &  &
    \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
     &  &
\end{flalign*}

@@

\paragraph{Robinson decomposition}
G-computation is one choice of decomposition of the CATE estimation into a sequence of regressions, where the only statistical estimate is the response function $\mu_{a}(x)$ \cite[chapter 15]{chernozhukov2024applied}. Other choices of decomposition exist, such as the R-decomposition \cite{robinson_rootnconsistent_1988}. It introduces two new statistical estimates, the conditional mean outcome
and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}):
\begin{flalign}
    \text{Conditional mean outcome} &                    & m(x) \myeq \; & \mathbb E_{Y \sim
    \mathcal{D}} [Y|X=x]            &                    &
    \label{def:m}
    \\
    \text{Propensity score}         &                    &
    e(x) \myeq \;                   & \mathbb P[A=1|X=x]
    \label{def:propensity_score}
\end{flalign}
the outcome can be written
\begin{flalign}\label{eq:r_decomposition}
    \text{R-decomposition}            &  & y(a) = m(x) + \big( a - e(x) \big)
    \tau(x) + \varepsilon(x; a)\notag &  &
    \\\ &&\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0&&
\end{flalign}
$m$ and $e$ are often called
\emph{nuisances} \cite{chernozhukov_double_2018}; they are unknown and are estimated in practice thanks to statistical models.

@@

We formalize model selection for causal estimation. Thanks to the outcome model identification (\autoref{eq:mu_identification}), a given model $f: \mathcal X
    \times \mathcal A \rightarrow \mathcal{Y}$ --learned from data or built from
domain knowledge-- induces feasible estimates of the ATE and CATE (eqs
\ref{eq:ate_estimate} and \ref{eq:cate_estimate}), $\hat \tau_{f}$ and $\hat \tau_{f}(x)$.
%
However, g-computation itself does not control error on both populations as highlighted in \ref{fig:toy_example}. Specific inductive biases of ML models are able to favor such regularization \cite{johansson2022generalization}. Thus, selection procedures, more elaborated than MSE are needed to favor such models. A selection procedure is defined by a risk $\ell$ evaluating the quality of a model $f$ on the observed data $O$ and a splitting strategy of the data.
%
Formally, let $\mathcal F=\{f: \mathcal X \times \mathcal A \rightarrow \mathcal{Y}\}$ be
a family of such estimators. Our goal is to select the best candidate in this
family for the observed dataset $O$ using a risk
$\ell$:
\begin{equation}
    f^*_{\ell} = \argmin_{f \in \mathcal{F}} \ell(f, O)
    \label{eq:causal_model_selection}
\end{equation}

@@

Using eight simulations setups from \cite{powers_methods_2018}, where
the oracle CATE is known, \cite{schuler_comparison_2018} compare four
causal risks, concluding that for CATE estimation the best
model-selection risk is the so-called $R\text{-risk}$
\cite{nie_quasioracle_2017} --def.\,\ref{def:r_risk}, below. Their
empirical results are clear for randomized treatment allocation but less
convincing for observational settings where both simple MSE --$\mu\text{-risk}(f)$ def.\,\ref{def:mu_risk}-- and
reweighted MSE --$\mu\text{-risk}_{IPW}$ def.\,\ref{def:mu_ipw_risk}--
appear to perform better than $R\text{-risk}$ on half of the simulations.
Another work \cite{alaa_validating_2019} studied empirically both MSE and
reweighted MSE risks on the semi-synthetic ACIC 2016 datasets
\cite{dorie_automated_2019}, but did not include the $R\text{-risk}$. We complete these
prior empirical work by studying a wider variety of data generative
processes and varying the influence of overlap, an important parameter of
the data generation process which makes a given causal metric appropriate
\cite{damour_overlap_2020}. We also study how to best adapt
cross-validation procedures to causal metrics which themselves come with
models to estimate.
